The Astropy Project: \\ Sustaining and Growing a Community-oriented Open-source Project and the Latest Major Release (v5.0) of the Core Package

### 2022-03-03 src/ms.tex
formatting, grammar, and clarity
Signed-off-by: nstarman <nstarkman@protonmail.com>

The \astropypkg core package is mature and stable, in that many features have
been part of the core package for years with (purposefully) few changes to the
software interface.
This maturity and stability allows for the broader astronomy \python community
to rely upon the \astropypkg core package and build specialized packages within
the Astropy ecosystem.
Even many non-astronomical \python libraries have come to rely on \astropypkg.

The relationship between \astropy and community is reciprocal, with the growing and evolving needs of the community and the maturation of the broader scientific \python ecosystem driving the development of the \astropypkg core package since \citealt{astropy:2018}.
% feels like ↑ could be expanded to 2 sentences

% Maybe briefly outline the sectors of  
% Using \astropypkg for
% precision astronomy has spurred development of faster and more accurate coordinate transformations and time calculations,
% super-computer work has spurred \astropypkg to be thread-safe and support \package{dask} \citep{daskpkg}
% Astropy has grown to interoperate, e.g. NumPy

Development may be roughly split into a few categories: new features added, performance enhancements, intra and inter-package interoperability, and improvements to precision, accuracy, and reproducibility.

% Summarize major pieces of development since the last paper (since v2.0). A few
% ideas (taken from What's New pages) below, but feel free to expand on or remove
% from this list:
% \begin{itemize}
%     \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
%     \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
%     \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
%     \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
%     \item APE 14 WCS
%     \item A new SpectralCoord class for representing and transforming spectral quantities
%     \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
% \end{itemize}

\subsubsection*{New Features} \label{sec:core-features-new}

\begin{itemize}
%  \item Time Series sub-package (V3.1: New Box Least Squares Periodogram)
%  \item Uncertainties and Distributions
  \item ERFA module is now it’s own thing; Obliquely tie in that units
    remains in astropy
  \item ? Pytest plugin ecosystem
%  \item A new SpectralCoord class for representing and transforming spectral quantities
%    Velocity in Coordinates
  \item WCS API : A shared Python interface for World Coordinate Systems
        Include APE reference
\end{itemize}

New features come in various forms, from new submodules providing
wholly new capabilities, to larger additions to existing submodules,
to large rewrites of the interface of modules.  This range is spanned
by the four main features that entered the astropy core, which we
describe below.

\paragraph{Uncertainties and Distributions.} Early in our carreers, we
were likely all taught that a measurement without a unit or an
uncertainty is essentially meaningless.  \astropyunits already
provides the ability to associate numbers with units, and propagate
these correctly, with its \astropyQuantity\ class, but so far the
ability to associate and propagate uncertainties is relatively
limited: in \astropysubpkg{nddata}, there are options to associate
errors with data arrays, but these are treated as independent, i.e.,
covariances are not tracked.

In general, proper error propagation is difficult, and often it is
best to use Monte Carle methods.  With the new
\astropysubpkg{uncertainties} sub-package, a first step is made to
make this possible.  It allows one to generate, for each variable,
randomly drawn samples in a \astropyDistribution, and then propagate
these by passing them through the normal analysis, producing a
\astropyDistribution\ of final results that can be inspected.

This \astropyDistribution\ can be any type of array, including of
course a \astropyQuantity, and the goal is to ensure it can be
seamlessly used to instantiate also other astropy classes, such as
\astropyskycoord\ and \astropyTime.

\paragraph{Time series.}  From sampling a continuous variable at fixed
times to counting events binned into time windows, many different
areas of astrophysics require the manipulation of time series.  These
are often logically tables, and the new \astropysubpkg{timeseries}
sub-package extends the \astropyQTable\ class to tables of data as a
function of time, where the data can either represent samples or
averages over particular time bins.  The new classes offer a number of
special methods to manipulate time series (folding, resampling, etc.)
and to read different data formats (such as {\em Kepler} lightcurves,
etc.).

@@

\paragraph{Spectral Coordinates.} Measurements are often taken at
specific ``spectral coordinates'', be they frequencies, wavelengths,
or photon energies.  The new \astropySpectralCoord\ class allows to
transform between these as well as to equivalent Doppler velocities.
It builds on the \astropyQuantity, which can also represent these
kinds of physical quantities, and allows conversion via dedicated
equivalencies, but aims to make this more straightforward.
Furthermore, it can be made aware of the observer and target reference
frames, allowing for instance transformation from telescope-centric (or
topocentric) frames to Barycentric or Local Standard of Rest
(LSRK and LSRD) velocity frames.

@@

@@

\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-performance}

\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-future}

>>>>

The \astropypkg core package is mature and stable, in that many features have
been part of the core package for years with (purposefully) few changes to the
software interface.  This maturity and stability allows for the broader
astronomy \python community to rely upon the \astropypkg core package and build
specialized packages within the Astropy ecosystem.  Even many non-astronomical
\python libraries have come to rely on \astropypkg. The relationship between
\astropy and the \python-using community is reciprocal, with the evolving needs
of the community and the maturation of the broader scientific \python ecosystem
driving much of the development of the \astropypkg core package.

Within \astropypkg, development may be roughly split into a few
categories: new features added, intra and inter-package interoperability,
improvements to precision, accuracy, and reproducibility, and performance
enhancements. We discuss these in turn.

Additionally, some features have been moved from the core package, either
because they are of more general use outside of \astropypkg, or because they are
too specialized and belong in an Astropy-affiliated package. An example of the
former is the copy of the IAU Standards Of Fundamental Astronomy (SOFA) software
Collection\footnote{\url{http://www.iausofa.org}} \citep{sofa} that \astropy
carried, as well as the \python wrappers for it. Since this is basic
infrastructure, with a release cycle set by SOFA, it made more sense to create
separate packages that serve as \astropypkg dependencies: \package{ERFA}
\citep{erfa} and \package{PyERFA} \citep{pyerfa}.  An example of something that
was too specialized was a Virtual Observatory sub-package; the Simple
Application Message Protocol made sense to keep (\astropysubpkg{samp}), but
other parts more properly belonged in \package{astroquery} (see
Sect.~\ref{sec:astroquery}).

\subsubsection*{New and Planned Features} \label{sec:core-features-new}

New features present a particular difficulty for a mature package such as
\astropypkg, because it is difficult to know {\em a priori} what the best
interface will be, and hard to address all possible use cases in one round of
development, even within \astropypkg\ itself.

New features come in various forms, from new submodules providing wholly new
capabilities, to larger additions to existing submodules, to large rewrites of
the interface of modules. This range is spanned by five main new features that
entered the \astropypkg core since the previous summary
\citepalias{astropy:2018}. In our descriptions of each we include how we hope
the new feature will evolve. Ultimately, in a user-driven project like \astropy,
development depends primarily on the priorities of contributors.

\paragraph{Uncertainties and Distributions} Early in our careers we were likely
all taught that measurements without units and uncertainties are meaningless.
\astropyunits provides the ability to associate numbers with units, and
propagate these correctly, using the \astropyQuantity\ class.  However, the
ability to associate and propagate uncertainties is relatively limited: in
\astropysubpkg{nddata}, there are options to associate errors with data arrays,
but these are treated as independent, i.e., covariances are not tracked.

In general, proper error propagation is difficult, and often it is best to use
Monte Carlo methods.  The new \astropysubpkg{uncertainties} sub-package is our
first step towards enabling seamless error propagation.  The package allows one
to generate, for each variable, randomly drawn samples in a
\astropyDistribution, and then propagate these by passing them through the
normal analysis, producing a \astropyDistribution\ of final results that can be
inspected.

This \astropyDistribution\ can be any type of array, including a
\astropyQuantity. The goal is to ensure \astropyDistribution\ can be seamlessly
used to instantiate other \astropypkg classes as well, such as \astropySkyCoord,
\astropyTime, and \astropyCosmology.

A plan for future development of \astropysubpkg{uncertainties} is to support
error propagation, tracking covariances, where the uncertainties are assumed to
be normally distributed rather than approximated by Monte-Carlo means. One
implementation problem to be solved is when to {\em stop} tracking covariances.
For instance, a simple operation such as subtracting the mean from $N$ data
points implies that all data points are now covariant with each other, i.e., one
has to carry $N\times N$ covariances.  For a large image, that becomes not just
pointless but also prohibitive in terms of memory and CPU usage.

\paragraph{Masked quantities}  It can be useful to mask bad data.  How one
supports masked data involves a number of choices, such as whether one indicates
data values are bad by a separate flag or by replacing them with a special
value, such as ``Not a Number'' (NaN).  In \astropypkg, both approaches have
been used: a flag for masked columns in \astropyTable, mask flags and bitmaps
for N-dimensional data in \astropysubpkg{nddata}, and replacing elements with
NaN in \astropyTime.  However masked quantities were poorly supported, making it
difficult to use masks in a \astropyQTable (in which \astropyQuantity\ is used
for all columns with units).

Unfortunately, the \texttt{MaskedArray} class from \package{numpy} could not be
used because its design made it a wrapper around any subclass. Therefore
another class could not use it as an alternative to a regular data array.
% Nathaniel: I find this topic sentence a bit confusing.
For quantities, \texttt{MaskedArray} does not work because a masked value does
not mean the unit is unknown. Consequently, a new \astropyMasked\ class was
designed, based on a framework very similar to that of \astropyDistribution,
making it easy to create masked instances of other classes.  In particular,
\astropyMasked\ can be used to create masked quantities for \astropyQTable,
enabling full support for reading and writing different file types.

The masked quantities work largely without any changes in higher-level objects
such as \astropySkyCoord, but work is still underway to expose the mask in those
classes, as well as to use masked arrays in \astropyTime\ instead of replacing
data with NaN.

\paragraph{Time series}  From sampling a continuous variable at fixed times to
counting events binned into time windows, many different areas of astrophysics
require the manipulation of time series.  The new \astropysubpkg{timeseries}
sub-package extends the \astropyQTable\ class to tables of data as a function of
time, where the data can either represent samples or averages over particular
time bins. The new classes offer a number of special methods to manipulate time
series (folding, resampling, etc.) and to read different data formats (such as
{\em Kepler} lightcurves, etc.).

@@

\paragraph{Spectral Coordinates} Measurements are often taken at specific
``spectral coordinates,'' be they frequencies, wavelengths, or photon energies.
\astropyQuantity\ can represent these physical quantities and convert between
them via dedicated equivalencies. The new \astropySpectralCoord\ builds on
\astropyQuantity\ by providing a more straightforward interface: baking in these
equivalencies and those for Doppler velocities. Furthermore,
\astropySpectralCoord\ can be made aware of the observer and target reference
frames, allowing transformation from telescope-centric (or topocentric) frames
to Barycentric or Local Standard of Rest (LSRK and LSRD) velocity frames.

\paragraph{A High-Level Interface to World Coordinate Systems}
Astronomical data are often provided alongside information about the
correspondence between ``real-world'' and pixel coordinates. This mapping is the
essence of the ``World Coordinate System (WCS)'' concept. From its inception,
the \astropysubpkg{wcs} sub-package allowed access to WCS information provided
in, e.g., FITS files, but it became clear that other WCS standards and
representations had to be supported for new missions and observatories (e.g.,
the James Webb Space Telescope and the Rubin Observatory).  To harmonize these
needs, a new high-level interface was created based on a formal design first
proposed in an \astropy\ Proposal for Enhancement (\astropyAPE{14}). The hope is
that by having a formal design, other packages implementing WCS objects can
straightforwardly modify their classes to conform to the new interface or build
thin wrappers that conform.  The implementation in \astropysubpkg{wcs} interacts
well with other \astropypkg\ objects such as \astropySkyCoord\ and \astropyTime.

@@

Because \astropy is modular and situated at the nexus between scientific
computing and astronomy, interoperability is an important focus for \astropy,
both within the various \astropypkg sub-packages and with other \python
libraries. If done right, all a person notices is that code using many different
python features and libraries ``just works.'' Here, we describe a few particular
efforts towards this goal.

\paragraph{\package{numpy} on Units}

\astropyQuantity\ is a backbone of \astropy, leveraging the power of
\package{numpy} and adding units. Previously, many \package{numpy} functions
would strip a \astropyQuantity\ of its units (or fail outright), limiting
\astropyQuantity's usefulness. Now, advances in \package{numpy} function
overloading (e.g.,
\href{https://numpy.org/neps/nep-0018-array-function-protocol.html}{NEP 18})
mean \astropyQuantity\ works with almost all \package{numpy} (v1.17+)
functions.

In the remaining gaps, the \package{numpy} and \astropyQuantity\
interoperability efforts are ongoing. For some functions, such as in
\href{https://numpy.org/doc/stable/user/basics.rec.html#module-numpy.lib.recfunctions}{\package{numpy}'s
module} for manipulating structured arrays, compatibility only requires
extending the existing \package{numpy}-\astropypkg bridge frameworks. Community
interest, in the form of a Feature Request (or better yet Pull Request) would be
sufficient to see this compatibility completed. For a few remaining functions,
discussed further in the
\href{https://docs.astropy.org/en/stable/known_issues.html#known-deficiencies}{\astropy
documentation}, the \package{numpy} framework does not yet allow for full
interoperability with \astropy. A goal of \astropy and the scientific \python
community is to enhance and implement the frameworks, allowing \astropyQuantity\
to propagate units seamlessly across the whole ecosystem of \package{numpy}-like
projects.

\paragraph{\astropypkg on Units}

Units were also integrated further within \astropypkg. For an already-defined unit-less
\astropyModel, e.g. those imported from another library, \astropymodeling can
now coerce units.
% TODO! transition words
A new module – \astropycosmologyunits – has been added to the cosmology
sub-package for defining and collecting cosmological units and equivalencies.
The unit \texttt{littleh} and equivalency
\astropyapidoc{cosmology.units.with_H0}{with\_H0} are deprecated from the main
\astropyunits sub-package and moved to \astropycosmologyunits. A new unit,
\texttt{redshift}, has been added for tracking factors of cosmological redshift. To
correctly use redshift units in dimensional analysis, \texttt{redshift} is
\astropyapidoc{cosmology.units.dimensionless_redshift}{treated as dimensionless}
by default. To convert between redshift and other cosmological distance
measures, e.g. CMB temperature or comoving distance, the equivalency
\astropyapidoc{cosmology.units.with_redshift}{with\_redshift}
has also been added. This equivalency is actually a composite of other equivalencies,
which may be used separately.

\paragraph{Table Mixin Columns}

Within \astropypkg and \python there are numerous ways to represent and store
array-valued data. Some of the differences are historical:
\astropyapidoc{table.Column}{table.Column} and
\astropyapidoc{io.fits.Column}{io.fits.Column} are not the same. Some
differences are computational: \package{numpy}, \package{cupy}, and
\package{dask} arrays are almost API identical, but are optimized for different
compute needs. Lastly, some differences are inherent: \astropyQuantity,
\astropyTime\, and \astropySkyCoord\ represent fundamentally different types of
objects.

\astropy believes that however array-valued data is stored, it should still be
usable as a column in a table. A well defined protocol for mixin-columns has
been developed for \astropytable and \astropyfits tables, allowing the original
object to be used as a column with a familiar API, and to ``round-trip'' through
tables with no loss of data or attributes. With this protocol, it is now
possible to store standard \astropyTableColumn\ objects to a FITS table
\href{https://docs.astropy.org/en/stable/io/fits/api/hdus.html}{HDU}. It is also
possible to store \astropy native objects -- including \astropyTime,
\astropyQuantity, and \astropySkyCoord --  within a \astropyTable. For objects
not already covered by the mixin protocol, functions can be registered with
\astropyTable\ to convert on-the-fly any array-like object into a mixin column.
As an example, the mixin functions are used to integrate \package{dask} arrays
with \astropyTable, allowing cloud-stored or cluster-scale data to be used as a
column in a \astropyTable.

\paragraph{Astropy FITS in Time}

The FITS standard was extended to rigorously describe time coordinates in the
World Coordinate System (WCS) framework \citep{FITS-Time:2015}. Compared to
other coordinates in the WCS framework, time requires more metadata: format,
scale, position, reference, etc. This metadata had to be manually specified and
the nuances understood by the user. \astropyfits could store this data in a
standard \astropyFitsColumn\ (for details, see the
\href{https://docs.astropy.org/en/stable/io/fits/index.html}{FITS
documentation}), but would not ``understand'' the metadata and attributes.
Thankfully, through the support of the
\href{https://summerofcode.withgoogle.com/archive/2017/projects/4778482366152704}{Google
Summer of Code} 2017 program, the \astropyfits package now allows \astropyTime\
as a `mixin' column. \astropyTime\ columns in FITS tables ``understand'' the
metadata while also providing a similar interface as the regular
\astropyFitsColumn\ objects. With \astropyTime\ columns, \astropy ensures that
time coordinates are stored according to the FITS standard in a consistent
manner and can be read/written without loss of information. For backward
compatibility with manual systems, this feature may be turned off, but it is
recommended to update. To learn more about this project, please see the final
report,
\href{https://aaryapatil.wordpress.com/2017/08/28/a-mixin-protocol-for-seamless-interoperability/}{A
mixin protocol for seamless interoperability}.

\paragraph{Persistent Storage}

\astropy's efforts to increase support for column types in a table is mirrored
by efforts to expand storage format options. New formats have been added, and
existing formats updated to support more column types.

\astropy now supports reading and writing tables in the American Astronomical
Society Journals' Machine-Readable Table (MRT) format. This ASCII format has
long been missing, and we are pleased that this year's Google Summer of Code
project adds MRT to \astropyTable\ I/O.  
For details, see the
\href{https://docs.astropy.org/en/stable/io/ascii/write.html#cds-mrt-format}{Machine-Readable
Table Format} section of \astropy's documentation. In addition, \astropyTable\ may also read from and write to
\href{https://asdf-standard.readthedocs.io/en/latest/}{ASDF},
\href{https://docs.astropy.org/en/stable/io/unified.html#table-io-parquet}{Parquet},
and \href{https://wwwastro.msfc.nasa.gov/qdp/}{QDP} formats.

The \href{https://github.com/astropy/astropy-APEs/blob/main/APE6.rst}{ECSV
standard} has been updated to version 1.0, adding support for reading and
writing \astropyTable\ with three additional data subtypes. First, it adds
multidimensional column data (both masked and unmasked) with fixed dimensions in
all table cells. Second, it adds multidimensional column data with
variable-dimension arrays similar to FITS variable-length arrays. Third, it adds
object-type columns with simple \python objects.

As mentioned previously, \package{dask} arrays may be used as a mixin column in
\astropyTable. Now, \package{dask} may also be the data array in
\astropyapidoc{io.fits.PrimaryHDU}{FITS HDU} and if written to disk, the array
will be computed while written, avoiding excessive memory use. \astropyTable\ can
now be appended to an existing FITS file, and \package{dask} mixin columns
interoperate seamlessly between the two table types.

\paragraph{Unified I/O architecture}

% TODO! better transition
\astropyiosubpkg{registry} is a powerful way to define input and output (I/O)
functions for \astropypkg objects, such as reading from or writing to a file.
\astropy uses this internally for the I/O methods in \astropyTable\ and
\astropyCosmology\ (I/O added v5.0+), and users can register custom I/O to extend
the options on these classes.

As of Astropy 5.0, the I/O registry submodule has switched to a class-based
architecture and has been refactored to allow read-only, write-only, and
read-write registries. Additionally, the registry submodule no longer assumes an
object has \astropy-style \texttt{read} and \texttt{write} methods when trying
to read or write that object. These two improvements enable a number of new use
cases for \astropyiosubpkg{registry}, which we hope have community adoption.
First, the class-based architecture allows for the creation of custom
registries. One application is in \astropycosmology, which has two different
\textit{kinds} of registries for \astropyCosmology: one for reading and writing
files, and another for converting between python objects. Second, \textit{any}
kind of object is now compatible with \astropyiosubpkg{registry}. Using
\astropyiosubpkg{registry} for custom classes means a developer does not have to
create a whole new I/O system and can offer to the user a unified and familiar
API.

@@

An important focus of \astropy has been to increase the precision and accuracy
of its functionality, while, where appropriate, making sure results obtained
using older versions of \astropypkg are reproducible. Within each section below
we include in our descriptions how old results may be reproduced. The primary
means are with configuration file options, and settings on runtime configuration
objects called \astropyScienceState. Notable improvements to \astropypkg have
been made to \astropytime, \astropyconstants, \astropycoordinates, and
\astropycosmology, and we describe each in turn.

\paragraph{Time}
In astronomy, time accuracy down to the (nano)second is frequently important.
Whether for planning observations, crunching pulsar or VLBI data, missing
seconds meaningfully impact results. Before, when using \astropyTime\ in
\astropytime leap seconds had to be manually applied; now, leap seconds are
applied automatically. Moreover, \astropypkg updates internal time data-files to
ensure that the correct leap second adjustment is always used. For
reproducibility, e.g. with an old code, \astropyLeapSeconds\ may also be
manually applied.

\paragraph{Constants}

Measurements of physical constants (and the units defined from them) improve
over time. Periodically the standardized systems of units and constants are
updated to reflect these improvements. For instance, in 2019 the SI system was
redefined \citep{NIST2019}, with an accompanying update to the physical
constants in
\href{https://codata.org/initiatives/data-science-and-stewardship/fundamental-physical-constants/}{\texttt{SI/CODATA
2018}}. \astropyconstants now defaults to use the \texttt{SI/CODATA 2018}
values, with the relation between units in \astropyunits based upon these
constants. Most of \astropypkg and affiliate packages build upon \astropyunits,
so this update has far-reaching effects across the \astropy ecosystem and
astronomy community.

For reproducibility, \astropypkg allows the constants' (and therefore units')
definitions to be rolled back to prior values, e.g. to the
\texttt{SI/CODATA 2014} values.
% CITATION
The constants' and units' definitions are set by configuration file as well as a
\astropyScienceState\. For work sensitive to the values of the fundamental
constants we recommend including with the work an \astropy configuration file,
which specifies the set of constants used.

\paragraph{Cosmology}
% TODO! At least one good sentence about "what is the benefit of
% astropy.cosmology"

\astropycosmology contains classes for representing cosmological models. Bundled
with the classes are commonly used \astropyCosmology\ realizations, e.g.,
best-fit measurements from
\href{https://lambda.gsfc.nasa.gov/product/map/current/}{WMAP}
\citep{WMAP2003} and
\href{https://www.nasa.gov/mission_pages/planck}{Planck}
\citep{PlanckMission:2006}.

When these important missions publish new measurements, \astropycosmology is
updated to include these results as realizations of the appropriate class
(generally \astropyFlatLambdaCDM). The models used by WMAP and Planck do not
always exactly correspond to \astropycosmology classes. For instance, the Planck
2018 results \citep{Planck2018VI:2020} include massive neutrinos in
$\Omega_{matter,0}$, while in \astropyFlatLambdaCDM\ this mass contribution is
stored in a separate parameter. Consequently, while some parameter values appear
different from the source paper, the \astropyCosmology\ realizations correctly
reproduce the cosmological models.

Since \citealt{astropy:2018} the following cosmology realizations have been
added:
\href{http://docs.astropy.org/en/stable/api/astropy.cosmology.WMAP1.html}{\texttt{WMAP1}}
from WMAP First Year \citep[Table 7: WMAP + CBI + ACBAR + 2dFGRS +
Lya]{WMAP1Year:2003};
\href{http://docs.astropy.org/en/stable/api/astropy.cosmology.WMAP3.html}{\texttt{WMAP3}}
from WMAP Three Year \citep[Table 6: WMAP + SNGold]{WMAP3Year:2007};
\href{http://docs.astropy.org/en/stable/api/astropy.cosmology.Planck15.html}{\texttt{Planck15}}
from Planck 2015 \citep[Table 4: TT, TE, EE + lowP + lensing +
ext]{Planck2015XIII:2016};
and \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.Planck18.html}{\texttt{Planck18}}
from the Planck 2018 best-fit cosmological parameters \citep[Table 2: TT, TE, EE
+ lowE + lensing + BAO]{Planck2018VI:2020}.

Astropy provides a
\href{http://docs.astropy.org/en/stable/api/astropy.cosmology.default_cosmology.html}{configurable
default cosmology}, which is used in calculations done in a cosmological
context. The default \astropyCosmology\ has been updated to the Planck 2018
parameters. For reproducibility, this cosmology may be set to old defaults, such
as the WMAP Three Year values. The default cosmology is dynamically configurable
using \astropyScienceState, meaning a set of calculations may be run using
different assumed cosmologies, and results compared between the two.

\paragraph{Coordinates}

Being able to describe the position of objects, i.e. their coordinates, is
fundamental across many knowledge domains. \astropypkg provides
\astropycoordinates for working with low-level positional and velocity data all
the way to high-level coordinate objects with reference frames, atmospheric
information, and more. A central feature of \astropycoordinates is the ability
to transform data between reference frames, e.g. \astropyICRS\ \citep{ICRS:1997}
to \astropyGalacticLSR\ \citep{GalacticLSR:2010}. However, there have been some
historical limitations for spatially proximate transformations, impacting the
usefulness of \astropycoordinates for ground-based telescopes and intra-solar
astronomy. Transformations in the \astropyAltAz\ frame were reasonably precise
for very distant objects, but wrong by up to several arcseconds for e.g., the
location of the moon. Now these transformations are much more precise, down to
the milliarcsecond level. Similar precision improvements were made to the
\astropyapi{coordinates.builtin_frames.HADec}{Hour Angle-Declination}
frame transformations. Additionally,
\astropyapi{coordinates.builtin_frames.BaseEclipticFrame}{Ecliptic frames}
and associated transformations have been updated to correctly reflect the “true”
(precession but not nutation) and “mean” (both precession and nutation)
terminology.

For Galactic astronomers, the \astropyGalactocentric\
\astropyapi{coordinates.galactocentric_frame_defaults}{frame defaults}
have been updated to include more recent measurements (tabulated in the
\href{https://docs.astropy.org/en/stable/api/astropy.coordinates.galactocentric_frame_defaults.html#astropy.coordinates.galactocentric_frame_defaults.references}{\texttt{frame\_defaults.references}}
attribute). For reproducibility, old definitions are still available, and will
continue to be so as \astropypkg is updated with forthcoming measurements of the
Galactic center, solar motion, etc.

\subsection*{Performance} \label{sec:core-features-performance}

As features are added and corner cases dealt with, performance often suffers.
Hence, pull requests that improve performance are welcomed, and were even the
main goal for one release cycle.  For most sub-packages, performance was
improved in the \python\ code, but for some, the most time-critical pieces were
rewritten in~C (e.g., for convolution of images, sigma-clipping, and converting
time strings to binary).  Furthermore, a particular effort was made to ensure
\astropypkg\ is thread-safe, so that it can be used on supercomputer clusters.
